---
title: "Coral Reef Health and Fish Occurrence Analysis Project"
author: "Jonah Kotzen, Elisabeth Laviolette, Lauren Geary, Elizabeth Breitmeyer,& Katie Miller,"
date: "2023-12-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rgbif)
library(tidyverse)
library(MuMIn)
library(rnoaa)
library(data.table)
library(ggmap)
library(usmap)
library(sf)
library(magick)
library(cowplot)
library(lme4)
library(car)
```

## Introduction
[Provide a brief overview of the project, its objectives, and its significance]

## Objectives
- Objective 1: [Detail the primary objective of your project]
- Objective 2: [Detail the secondary objective of your project, if any]

## Methodology
Briefly describe the methodology used in this project, including data sources and analysis techniques.

### Data Sources
- Coral Reef Health Data: [Describe the source and nature of the coral reef health data]
- Fish Occurrence Data: Using the `rgbif` package to access data from GBIF.

### Data Retrieval and Processing
#### Using `rgbif` for Fish Occurrence Data
```{r querying GBIFs API, echo=FALSE, cache = TRUE}
species <- c("", "", "", "","")
year <- paste0("2000",",","2022")

data.l <- list() #making empty list to store data

##Results

##Discussion
##Author Contributions
##References
